---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
payments <- clean_names(payments)

# had to get chat gpt to help me out with these
school_name_finder <- payments %>%
  filter(str_detect(vendor_name, "SCHOOL")) %>%
  group_by(vendor_name) %>%
  summarise(count = n())

pg_schools <- payments %>%
  mutate(vendor_name = str_to_upper(vendor_name)) %>%
  filter(str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC") |
         str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL") |
         str_detect(vendor_name, "PG COUNTY PUBLIC"))

nrow(pg_schools)
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1:

```{r}

school_name_finder <- payments |> filter(str_detect(vendor_name,"SCHOOL")) |> group_by(vendor_name) |> summarize(count=n())

pg_schools <- payments |>
  filter(str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC") | str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL") | str_detect(vendor_name, "PG COUNTY PUBLIC"))
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2:

```{r}

#also needed some help from chat gpt on this one but it makes sense. grabbing from pg schools we get pg totals which is grouped by fiscal year and a summation of the total amount which is the sum of all the pg school payments
pg_totals <- pg_schools %>%
  group_by(fiscal_year) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

pg_totals

pg_totals %>% filter(total_amount == max(total_amount))
```
1470133533

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3.

```{r}
#i had to use chat gpt again for this and even it couldnt figure it out. I'm a little confused as tob why it wont find what im asking for?

umd_capital <- payments %>%
  filter(str_detect(str_to_upper(description), "Capital") &
         str_detect(str_to_upper(vendor_name), "University"))

vendor_totals <- umd_capital %>%
  group_by(vendor_name, vendor_zip) %>%
  summarise(total_amount = sum(amount, na.rm=TRUE)) %>%
  arrange(desc(total_amount))

vendor_totals
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4:

```{r}

md_grants_loans <- janitor::clean_names(md_grants_loans)

stemcell_grants <- md_grants_loans %>%
  filter(str_detect(str_to_lower(description), "stem cell"))

#I dont understand why it wont find anything

stemcell_totals <- stemcell_grants %>%
  group_by(fiscal_year) %>%
  summarise(total_amount = sum(amount, na.rm=TRUE)) %>%
  arrange(fiscal_year)

stemcell_totals
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5:

```{r}
#I had to use chat gpt again to figure thus one out but i think i understand what the funcitons are doing
md_grants_loans <- clean_names(md_grants_loans)


umd_zip <- "20742"

college_park_grantees <- md_grants_loans %>%
  filter(zip_code == "20742") %>%
  group_by(grantee) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))

college_park_grantees
```
